---
title: "Looking at FragPipe TMT ouptuts"
output: 
  pdf_document: 
    toc: yes
date: "2022-12-15"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 6, fig.height = 6)
```


# annotation.tsv

```{r}
library(tidyverse)
library(prolfqua)

FP = 19
if ( FP == 18 ) {
  sampleCol <- "sample-"
  input_folder <- "../inst/extdata/FragPipe18_GUI_TMT10/"
  
} else {
  sampleCol <- "sample_"
  input_folder <- "../inst/extdata/FragPipe19_GUI_TMT10/"
  
}

print(input_folder)
annot <- read.table(file.path(input_folder, "annotation_annotation.txt"),header = FALSE, sep = " " )
colnames(annot) <- c("channel", "sample")
annot <- annot |> dplyr::mutate(channel = paste0("channel_", channel))

A = c("channel_126", "channel_127N", "channel_127C", "channel_128N", "channel_128C")
annot <- annot |> mutate(group = case_when(channel %in% A ~ "A", TRUE ~ "B"))
annot <- annot |> unite("sampleName", group, channel, sep = "_", remove=FALSE)

```



# Philosopher ouptuts

## protein.tsv




```{r pressure, echo=FALSE}

prot <- read_tsv(file.path(input_folder,"protein.tsv"))
prot <- prot |> select(all_of(c("Protein")),starts_with(sampleCol))
protl <- prot |> pivot_longer(starts_with(sampleCol), values_to = "abundance", names_to = "sample")
protl <- dplyr::inner_join(annot , protl )

```

```{r}

table <- prolfqua::AnalysisTableAnnotation$new()
table$hierarchy["Protein"] <- "Protein"
table$factors["group"] = "group"
table$fileName = "sample"

table$set_response("abundance")
config <- prolfqua::AnalysisConfiguration$new(table)
data <- prolfqua::setup_analysis(protl, config)
ldata <- LFQData$new(data, config)
ldata$hierarchy_counts()

abd <- ldata$data |> group_by(sampleName) |> summarize(sum_abundance = sum(abundance))
```

```{r fig.cap = "Barplot showing Total abundances per file for protein.tsv"}

abd |> ggplot(aes(x = sampleName, y = sum_abundance)) + 
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white" ) + 
  geom_text(aes(label = round(sum_abundance,digits= 0)), vjust=-0.3, size=3.5 , angle = 90) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


```{r fig.cap = "Density of log2 transformed data."}
ldata$remove_small_intensities()
pl <- ldata$get_Transformer()$log2()$lfq$get_Plotter()
pl$intensity_distribution_density()

```

```{r fig.cap = "PCA plot."}
pl$pca()


hm <- pl$heatmap()

```


```{r, fig.cap = "Heatmap log2 trasformed data and row scaling."}
hm
```


\pagebreak


## peptide.tsv


```{r look_at_peptide_tsv}
peptides <- read_tsv(file = file.path(input_folder,"peptide.tsv"))
peptides <- peptides |> dplyr::select(
  dplyr::all_of(c("Peptide", "Protein", "Protein ID",   "Entry Name",   "Gene", "Protein Description")),
  dplyr::starts_with(sampleCol))

peptides <- peptides |> tidyr::pivot_longer(dplyr::starts_with(sampleCol), names_to = "sample", values_to = "abundance")


if (FP == 18) {
  peptides <- peptides |> mutate(channel = gsub("^sample\\-01", "channel_", sample))
  annot$sample <- NULL
  peptides <- dplyr::inner_join(annot , peptides, by = "channel" )
} else{
  peptides$channel |> unique()
  
  peptides <- peptides |> mutate(channel = gsub("^sample\\.01", "channel_", sample))
  peptides <- dplyr::inner_join(annot , peptides, by = c("sample" =  "channel" ))

}


```



```{r}

table <- prolfqua::AnalysisTableAnnotation$new()
table$hierarchy["Protein"] <- "Protein"
table$hierarchy["Peptide"] <- "Peptide"
table$factors["group"] = "group"
table$fileName = "channel"
table$sampleName = "sampleName"
table$set_response("abundance")
config <- prolfqua::AnalysisConfiguration$new(table)
data <- prolfqua::setup_analysis(peptides, config)
ldata <- LFQData$new(data, config)
ldata$hierarchy_counts()

abd <- ldata$data |> dplyr::group_by(sampleName) |> dplyr::summarize(sum_abundance = sum(abundance))


abd |> ggplot(aes(x = sampleName, y = sum_abundance)) + 
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white" ) + 
  geom_text(aes(label = round(sum_abundance,digits= 0)), vjust=-0.3, size=3.5 , angle = 90) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```



```{r, fig.cap = "Intensity distribution log2 transformed peptide abundances."}
ldata$remove_small_intensities()
pl <- ldata$get_Transformer()$log2()$lfq$get_Plotter()

pl$intensity_distribution_density()

```

```{r fig.cap = "PCA plot using peptide abundances (log2)."}
pl$pca()
hm <- pl$heatmap()

```


```{r fig.cap = "Heatmap log2 peptide abundances."}

hm
```

\pagebreak

## estimate protein intensities using median polish

- protein intensities are estimates from log2 peptide abundances using tukeys median polish
- protein intensities are centered and scaled

```{r}
ldata$remove_small_intensities()
ldata$hierarchy_counts()
#tmp <- ldata$filter_proteins_by_peptide_count()
#ldata$hierarchy_counts()

llog2 <- ldata$get_Transformer()$log2()$lfq
agg <- llog2$get_Aggregator()
agg$medpolish()
protl <- agg$lfq_agg
```

```{r}
protl$hierarchy_counts()
protl <- protl$get_Transformer()$robscale()$lfq

pl <- protl$get_Plotter()
```

```{r fig.cap = "Intensity distribution protein abundances estimated using median polish."}
pl$intensity_distribution_density()
```

```{r fig.cap = "PCA plot log2 and scaled protein abundances."}
pl$pca()
naheat <- pl$NA_heatmap()
```

```{r fig.cap = "Missing values heatmap - protein."}
naheat
hm <- pl$heatmap()
```

```{r fig.cap = "Hatmap protein abundances log2."}
hm
```

\pagebreak

## Compute fold changes between group A and B

```{r}

model <- strategy_lm( paste0(protl$response(), "~ group"))

Ct <- c("AvsB" = "groupA - groupB")
tmp <- build_model(protl, model)
ctr <- Contrasts$new(tmp, Ct)
tmp <- ctr$get_Plotter()
tmp$volcano()$FDR
```



```{r}
dd <- ctr$get_contrasts()

dd <- dd |> filter(!grepl("contam_", Protein)) 
dd <- dd |> separate(Protein, c(NA, "species") , sep = "_", remove = FALSE) 
dd <- dd |> filter(species %in% c("HUMAN", "ECOLI", "SOYBN"))

ggplot(dd, aes(x = species, y = diff)) + geom_violin() + geom_boxplot(width = 0.07)

```

```{r}
ggplot(dd, aes(x = avgAbd, y = diff, color = species)) + geom_point(alpha = 0.6) + theme_bw()

```


```{r}

dd |> group_by(species) |>
  dplyr::summarise(
    n = n(), 
    median = median(diff),
    mean = mean(diff),
    min = min(diff),
    max = max(diff),
    quant25 = quantile(diff, 0.25),
    quant75 = quantile(diff, 0.75),
    interquartileRange = quant75 - quant25  ) |> knitr::kable(digits = 2)

```



```{r}
par(mfrow = c(1,2))
dd <- dd |> mutate(TP = (! species %in% c("HUMAN","ECOLI") ))
pr <- pROC::roc(dd$TP, dd$statistic, partial.auc = c(1,0.9))
plot(pr, main = "SOYBN")
print("SOYBN")
print(pr)

dd <- dd |> mutate(TP = (! species %in% c("HUMAN","SOYBN") ))
pr <- pROC::roc(dd$TP, dd$statistic, partial.auc = c(1,0.9))
plot(pr, main = "ECOLI")
print("ECOLI")
print(pr)

```



\pagebreak

# TMT- integrator

## Protein tmt-integrator None

```{r}
protNone <- read_tsv(file.path(input_folder,"/tmt-report/abundance_protein_None.tsv"))
#head(protNone)
proteinNone_long <- pivot_longer(protNone, starts_with(sampleCol), values_to = "abundance", names_to = "channel")
#head(proteinNone_long)
protTMT_integrator <- inner_join(annot, proteinNone_long, by=c("sample" = "channel"))
#head(protTMT_integrator)
```


```{r}
table <- prolfqua::AnalysisTableAnnotation$new()
table$hierarchy["Protein"] <- "Index"
table$factors["group"] = "group"
table$fileName = "sample"
table$set_response("abundance")
config <- prolfqua::AnalysisConfiguration$new(table)

data <- prolfqua::setup_analysis(protTMT_integrator, config)
ldata <- LFQData$new(data, config)
ldata$hierarchy_counts()
#ldata$remove_small_intensities(threshold = 15)
#ldata$hierarchy_counts()

abd <- ldata$data |> group_by(sampleName) |> summarize(sum_abundance = sum(abundance, na.rm = TRUE))
abd |> ggplot(aes(x = sampleName, y = sum_abundance)) + 
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white" ) + 
  geom_text(aes(label = round(sum_abundance,digits= 0)), vjust=-0.3, size=3.5 , angle = 90) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r fig.cap = "intensity distribution tmt-integrator none."}
pl <- ldata$get_Plotter()
pl$intensity_distribution_density()

hm <- pl$heatmap()
```



```{r fig.cap = "heatmap tmt-integrator None"}
hm
```

\pagebreak

## Protein tmt-integrator MD


```{r}
protNone <- read_tsv(file.path(input_folder,"/tmt-report/abundance_protein_MD.tsv"))
proteinNone_long <- pivot_longer(protNone, starts_with(sampleCol), values_to = "abundance", names_to = "channel")
protTMT_integrator <- inner_join(annot, proteinNone_long, by=c("sample" = "channel"))
```

```{r}
fasta <- prozor::readPeptideFasta("../inst/extdata/FragPipe18_GUI_TMT10/Hum-Ecoli-Soy-finalDB-22k.fasta")
ds <- data.frame(ids = names(fasta))
ds <- ds |> separate(ids, c(NA, "ID", "Name"), sep = "\\|", remove = FALSE) 
protTMT <- inner_join(ds , protTMT_integrator,  by = c("ID"=  "Index"))

```

```{r}
table <- prolfqua::AnalysisTableAnnotation$new()
table$hierarchy["Protein"] <- "ids"
table$factors["group"] = "group"
table$fileName = "sample"
table$set_response("abundance")
config <- prolfqua::AnalysisConfiguration$new(table)

data <- prolfqua::setup_analysis(protTMT, config)
ldata <- LFQData$new(data, config)
ldata$hierarchy_counts()
```


```{r}
abd <- ldata$data |> group_by(sampleName) |> 
  summarize(sum_abundance = sum(abundance, na.rm = TRUE), median_abd = median(abundance, na.rm =TRUE), sd = sd(abundance - median(abundance, na.rm =TRUE), na.rm = TRUE))
knitr::kable(abd, digits = 2)
```

```{r}
abd |> ggplot(aes(x = sampleName, y = sum_abundance)) + 
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white" ) + 
  geom_text(aes(label = round(sum_abundance,digits= 0)), vjust=-0.3, size=3.5 , angle = 90) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
pl <- ldata$get_Plotter()
pl$pca()
```

```{r}
pl$intensity_distribution_density()
hm <- pl$heatmap()
```



```{r}
hm
```


\pagebreak

### fit model



```{r}
model <- strategy_lm( paste0(ldata$response(), "~ group"))

Ct <- c("AvsB" = "groupA - groupB")
tmp <- build_model(ldata, model)
ctr <- Contrasts$new(tmp, Ct)
dd <- ctr$get_contrasts()

tmp <- ctr$get_Plotter()
tmp$volcano()$FDR
```



```{r}
dd <- dd |> filter(!grepl("contam_", Protein)) 

dd <- dd |> separate(Protein, c(NA, "species") , sep = "_", remove = FALSE) 
ggplot(dd, aes(x = species, y = diff)) + geom_violin() + geom_boxplot(width = 0.07)
```


```{r}
ggplot(dd, aes(x = avgAbd, y = diff, color = species)) + geom_point(alpha = 0.6) + theme_bw()
ggplot(dd, aes(x = rank(avgAbd), y = diff, color = species)) + geom_point(alpha = 0.6) + theme_bw()

```


```{r}
dd |> group_by(species) |>
  summarise(
    n = n(), 
    median = median(diff),
    mean = mean(diff),
    min = min(diff),
    max = max(diff),
    quant25 = quantile(diff, 0.25),
    quant75 = quantile(diff, 0.75),
    interquartileRange = quant75 - quant25  ) |> knitr::kable(digits = 2)

```


```{r}
par(mfrow = c(1,2))
dd <- dd |> mutate(TP = (! species %in% c("HUMAN","ECOLI") ))
pr <- pROC::roc(dd$TP, dd$statistic, partial.auc = c(1,0.9))
plot(pr, main = "SOYBN")
print("SOYBN")
print(pr)

dd <- dd |> mutate(TP = (! species %in% c("HUMAN","SOYBN") ))
pr <- pROC::roc(dd$TP, dd$statistic, partial.auc = c(1,0.9))
plot(pr, main = "ECOLI")
print("ECOLI")
print(pr)

```


\pagebreak

## Protein tmt-integrator GN

```{r}
protNone <- read_tsv(file.path(input_folder,"/tmt-report/abundance_protein_GN.tsv"))
#head(protNone)
proteinNone_long <- pivot_longer(protNone, starts_with(sampleCol), values_to = "abundance", names_to = "channel")
#head(proteinNone_long)
protTMT_integrator <- inner_join(annot, proteinNone_long, by=c("sample" = "channel"))
#head(protTMT_integrator)
```


```{r}
table <- prolfqua::AnalysisTableAnnotation$new()
table$hierarchy["Protein"] <- "Index"
table$factors["group"] = "group"
table$fileName = "sample"
table$set_response("abundance")
config <- prolfqua::AnalysisConfiguration$new(table)

data <- prolfqua::setup_analysis(protTMT_integrator, config)
ldata <- LFQData$new(data, config)
ldata$hierarchy_counts()

```


```{r}
abd <- ldata$data |> group_by(sampleName) |> summarize(sum_abundance = sum(abundance, na.rm = TRUE), median_abd = median(abundance, na.rm =TRUE), sd = sd(abundance - median(abundance, na.rm =TRUE), na.rm = TRUE))
knitr::kable(abd, digits = 2)
```


```{r}

abd |> ggplot(aes(x = sampleName, y = sum_abundance)) + 
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white" ) + 
  geom_text(aes(label = round(sum_abundance,digits= 0)), vjust=-0.3, size=3.5 , angle = 90) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pl <- ldata$get_Plotter()
```

```{r}
pl$pca()
```

```{r}
pl$intensity_distribution_density()
hm <- pl$heatmap()
```

```{r}
hm
```







